Pipeline Performance Figure Generation

### 2023-05-30 TestingPipeline/Scripts/paper_figure_generation_extended.Rmd
Difference of adding +0.5 rules.

Function that allows for the comparing pieces of the pipeline
```{r getting_viral_set_1}
getting_viral_set_1 <- function(input_seqs,
                                include_vibrant=FALSE, 
                                include_virsorter2=FALSE,
                                include_deepvirfinder=FALSE,
                                include_tuning_viral=FALSE,
                                tv_1=T, tv_2=T, tv_3=T, tv_4=T,
                                include_tuning_not_viral=FALSE,
                                ntv_1=T, ntv_2=T, ntv_3=T, ntv_4=T, ntv_5=F,
                                include_virsorter=FALSE) {

@@

  if (include_vibrant) {
    keep_score[input_seqs$vibrant_quality=="high quality draft"] <- keep_score[input_seqs$vibrant_quality=="high quality draft"] + 1
    keep_score[input_seqs$vibrant_quality=="medium quality draft"] <- keep_score[input_seqs$vibrant_quality=="medium quality draft"] + 1
    keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$provirus] <- keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$provirus] + 0.5
  }

  if (include_virsorter2) {
    keep_score[input_seqs$max_score>=0.50] <- keep_score[input_seqs$max_score>=0.50] + 0.5
    keep_score[input_seqs$max_score>=0.95] <- keep_score[input_seqs$max_score>=0.95] + 0.5  
  }

  if (include_virsorter) {
    keep_score[input_seqs$category==1] <- keep_score[input_seqs$category==1] + 1
    keep_score[input_seqs$category==2] <- keep_score[input_seqs$category==2] + 0.5
    keep_score[input_seqs$category==3] <- keep_score[input_seqs$category==3] + 0.5
    keep_score[input_seqs$category==4] <- keep_score[input_seqs$category==4] + 1
    keep_score[input_seqs$category==5] <- keep_score[input_seqs$category==5] + 0.5 
    keep_score[input_seqs$category==6] <- keep_score[input_seqs$category==6] + 0.5 
  }

  if (include_deepvirfinder) {
     # add if DVF calls viral
    keep_score[(input_seqs$score>=0.9 & input_seqs$pvalue<=0.05) & input_seqs$checkv_length<20000] <- keep_score[(input_seqs$score>=0.9 & input_seqs$pvalue<=0.05) & input_seqs$checkv_length<20000] + 0.5
    keep_score[(input_seqs$score>=0.7 & input_seqs$pvalue<=0.05) & input_seqs$checkv_length<20000] <- keep_score[(input_seqs$score>=0.7 & input_seqs$pvalue<=0.05) & input_seqs$checkv_length<20000] + 0.5
  }

@@

this function builds a list of all of the combinations that the user wants to 
test. 
In this case, we're comparing the performance of all unique combinations of the 
six tools.
```{r}
build_score_list <- function(input_seqs, combos) {
  output <- data.frame(precision=rep(0, nrow(combos)),
                       recall=rep(0, nrow(combos)),
                       F1=rep(0, nrow(combos)),
                       MCC=rep(0, nrow(combos)),
                       prop_viral=rep(0, nrow(combos)))
  for (i in 1:nrow(combos)) {
    keep_score <- getting_viral_set_1(input_seqs, include_vibrant = combos$VIBRANT[i],
                                            include_virsorter = combos$VS[i],
                                            include_virsorter2 = combos$VS2[i],
                                            include_tuning_viral = combos$tune_viral[i],
                                            include_tuning_not_viral = combos$tune_not_viral[i],
                                            include_deepvirfinder = combos$DVF[i])

@@

## Calculate the performance of each pipeline
quantifying variability between rule sets
```{r}
accuracy_scores <- data.frame(testing_set_index=rep(0, nrow(combos_list)*8),
                      precision=rep(0, nrow(combos_list)*8),
                       recall=rep(0, nrow(combos_list)*8),
                       F1=rep(0, nrow(combos_list)*8),
                       MCC=rep(0, nrow(combos_list)*8), 
                      prop_viral=rep(0, nrow(combos_list)*8))

accuracy_scores <- cbind(testing_set_index=rep(1, nrow(combos_list)),
                              build_score_list(viruses[viruses$Index==1,], combos_list))
for (i in 2:8) {
  accuracy_scores <- rbind(accuracy_scores,
                           cbind(testing_set_index=rep(i, nrow(combos_list)),
                              build_score_list(viruses[viruses$Index==i,], combos_list)))
}
```

@@

```{r}
MCC_sd_quant <- cbind(accuracy_scores$toolcombo[accuracy_scores$testing_set_index==1], as.data.frame(matrix(data=0, nrow=63, ncol=7)))

for (i in (2:8)) {
  temp <- accuracy_scores %>%
    filter(testing_set_index<=i) %>%
    select(MCC, toolcombo) %>%
    group_by(toolcombo) %>%
    summarise(sd = sd(MCC))

  MCC_sd_quant[,i] <- temp$sd
}

colnames(MCC_sd_quant) <- c("testing_set_index", "1-2", "1-3", "1-4",
                            "1-5", "1-6", "1-7", "1-8")

MCC_sd_quant <- MCC_sd_quant %>% pivot_longer(cols=contains("-"))

ggplot(MCC_sd_quant, aes(x=testing_set_index, y=value, 
                                  color=name, fill=name)) +
  geom_point(alpha=0.5)

ggplot(MCC_sd_quant, aes(x=name, y=value)) +
  geom_boxplot(alpha=0.5, color=pal(4)[1]) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 0),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Testing Sets") +
  ylab("Standard Deviation") 
```

## Calculate the performance of each pipeline
```{r}
accuracy_scores <- accuracy_scores %>% filter(testing_set_index<=5)
```

@@

```{r}
ggsave(
  "../IntermediaryFiles/simplified_performance_metricrule_types.png",
  plot = fig,
  scale = 1,
  width = 6.5,
  height = 4,
  units = c("in"),
  dpi = 300,
  limitsize = TRUE
)
```

visualizing change when add the tnv rule
```{r}
accuracy_scores_one_ts <- accuracy_scores[accuracy_scores$testing_set_index==1,]

tnv <- rep(NA, 63)

for (i in 1:nrow(accuracy_scores_one_ts)) {
  tnv[i] <- as.numeric(substr(accuracy_scores_one_ts$toolcombo[i], 5, 5))
}

num_no_tnv <- 63-(sum(tnv))
```

```{r}
no_tnv_prop <- data.frame(tool_combo=accuracy_scores_one_ts$toolcombo[grep("0", tnv)],
                          prop_viral=rep(0, num_no_tnv),
                          MCC=rep(0, num_no_tnv),
                          precision=rep(0, num_no_tnv),
                          recall=rep(0, num_no_tnv))

tc_tnv <- "0"
tc_no_tnv <- "0"
prop_tnv <- 0
j <- 1

for (i in grep("0", tnv)) {
  tc_no_tnv <- as.character(accuracy_scores_one_ts$toolcombo[i])
  tc_tnv <- tc_no_tnv
  substr(tc_tnv, 5, 5) <- "1"

  prop_tnv <- accuracy_scores_one_ts$prop_viral[grep(tc_tnv,accuracy_scores_one_ts$toolcombo)][1]
  no_tnv_prop$prop_viral[j] <-
    (prop_tnv-accuracy_scores_one_ts$prop_viral[i])/prop_tnv*100

  prop_tnv <- accuracy_scores_one_ts$MCC[grep(tc_tnv,accuracy_scores_one_ts$toolcombo)][1]
  no_tnv_prop$MCC[j] <-
    (prop_tnv-accuracy_scores_one_ts$MCC[i])/prop_tnv*100

  prop_tnv <- accuracy_scores_one_ts$precision[grep(tc_tnv,accuracy_scores_one_ts$toolcombo)][1]
  no_tnv_prop$precision[j] <-
    (prop_tnv-accuracy_scores_one_ts$precision[i])/prop_tnv*100

  prop_tnv <- accuracy_scores_one_ts$recall[grep(tc_tnv,accuracy_scores_one_ts$toolcombo)][1]
  no_tnv_prop$recall[j] <-
    (prop_tnv-accuracy_scores_one_ts$recall[i])/prop_tnv*100

  j <- j+1
}

mean(no_tnv_prop$prop_viral)
sd(no_tnv_prop$prop_viral)

no_tnv_prop_melt <- no_tnv_prop %>% 
  select(precision, recall, MCC, prop_viral, tool_combo) %>%
  pivot_longer(cols=c(precision, recall, MCC, prop_viral), 
               names_to="performance_metric",
               values_to="performance_metric_percent_diff")

visualizing when add the tuning addition rule
```{r}
accuracy_scores_one_ts <- accuracy_scores[accuracy_scores$testing_set_index==1,]

tv <- rep(NA, 63)

for (i in 1:nrow(accuracy_scores_one_ts)) {
  tv[i] <- as.numeric(substr(accuracy_scores_one_ts$toolcombo[i], 1, 1))
}

num_no_tv <- 63-(sum(tv))

no_tv_prop <- data.frame(tool_combo=accuracy_scores_one_ts$toolcombo[grep("0", tv)],
                          prop_viral=rep(0, num_no_tv),
                          MCC=rep(0, num_no_tv),
                          precision=rep(0, num_no_tv),
                          recall=rep(0, num_no_tv))

tc_tv <- "0"
tc_no_tv <- "0"
prop_tv <- 0
j <- 1

for (i in grep("0", tv)) {
  tc_no_tv <- as.character(accuracy_scores_one_ts$toolcombo[i])
  tc_tv <- tc_no_tv
  substr(tc_tv, 1, 1) <- "1"

  prop_tv <- accuracy_scores_one_ts$prop_viral[grep(tc_tv,accuracy_scores_one_ts$toolcombo)][1]
  no_tv_prop$prop_viral[j] <-
    (prop_tv-accuracy_scores_one_ts$prop_viral[i])/prop_tv*100

  prop_tv <- accuracy_scores_one_ts$MCC[grep(tc_tv,accuracy_scores_one_ts$toolcombo)][1]
  no_tv_prop$MCC[j] <-
    (prop_tv-accuracy_scores_one_ts$MCC[i])/prop_tv*100

  prop_tv <- accuracy_scores_one_ts$precision[grep(tc_tv,accuracy_scores_one_ts$toolcombo)][1]
  no_tv_prop$precision[j] <-
    (prop_tv-accuracy_scores_one_ts$precision[i])/prop_tv*100

  prop_tv <- accuracy_scores_one_ts$recall[grep(tc_tv,accuracy_scores_one_ts$toolcombo)][1]
  no_tv_prop$recall[j] <-
    abs(prop_tv-accuracy_scores_one_ts$recall[i])/prop_tv*100

  j <- j+1
}

mean(no_tv_prop$prop_viral)
sd(no_tv_prop$prop_viral)

no_tv_prop_melt <- no_tv_prop %>% 
  select(precision, recall, MCC, prop_viral, tool_combo) %>%
  pivot_longer(cols=c(precision, recall, MCC, prop_viral), 
               names_to="performance_metric",
               values_to="performance_metric_percent_diff")
```

```{r}
no_tv_and_ntv_prop_melt <- rbind(no_tv_prop_melt, no_tnv_prop_melt)

no_tv_and_ntv_prop_melt$set <- c(rep("tuning addition", nrow(no_tv_prop_melt)),
                                 rep("tuning removal", nrow(no_tnv_prop_melt)))

pal <- ggthemes::tableau_color_pal(palette="Tableau 10", type="regular")

fig <- ggplot(no_tv_and_ntv_prop_melt, 
              aes(x=performance_metric, y=performance_metric_percent_diff,
                  color=set, fill=set)) +
  geom_boxplot() +
  #geom_point() +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=10),
    axis.text.x=element_text(size=12),
    legend.text=element_text(size=12),
    axis.title=element_text(size=12),
    strip.background = element_rect(fill="white", color="grey"),
    strip.text = element_text(color="black", size=12)
  ) +
  xlab("Performance Metric") +
  ylab("Percent Difference") +
  geom_hline(yintercept = 1, color="grey") +
  scale_fill_manual(name="",
                     values = alpha(c(pal(6)[c(6,1)]), 0.4)) +
  scale_color_manual(name="",
                     values = alpha(c(pal(6)[c(6,1)]), 0.6))

@@

ggsave(
  "../IntermediaryFiles/all_MCC_scatterplot.png",
  plot = fig,
  scale = 1,
  width = 8,
  height = 4,
  units = c("in", "cm", "mm", "px"),
  dpi = 300,
  limitsize = TRUE
)

ggplot(accuracy_scores, aes(x=toolcombo, y=MCC, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=20),
  ) +
  xlab("Tool Combination (tv, DVF, tnv, VB, VS, VS2)") +
  ylab("MCC") +
  scale_fill_manual(name="Number of Tools",
                     values = alpha(rev(viridis(6)), 0.5)) +
  scale_color_manual(name="Number of Tools",
                     values = alpha(rev(viridis(6)), 1))
```
```{r}
max(accuracy_scores$MCC)
```
=78

@@

>>>>

Function that allows for the comparing pieces of the pipeline

```{r getting_viral_set_1}
getting_viral_set_2 <- function(input_seqs,
                                include_vibrant=FALSE, 
                                include_virsorter2=FALSE,
                                include_deepvirfinder=FALSE,
                                include_tuning_viral=FALSE,
                                tv_1=T, tv_2=T, tv_3=T, tv_4=T,
                                include_tuning_not_viral=FALSE,
                                ntv_1=T, ntv_2=T, ntv_3=T, ntv_4=T, ntv_5=F,
                                include_virsorter=FALSE) {

@@

  if (include_vibrant) {
    keep_score[input_seqs$vibrant_quality=="high quality draft"] <- keep_score[input_seqs$vibrant_quality=="high quality draft"] + 1
    keep_score[input_seqs$vibrant_quality=="medium quality draft"] <- keep_score[input_seqs$vibrant_quality=="medium quality draft"] + 1
    #keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$provirus] <- keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$provirus] + 0.5
  }

  if (include_virsorter2) {
   # keep_score[input_seqs$max_score>=0.50] <- keep_score[input_seqs$max_score>=0.50] + 0.5
    keep_score[input_seqs$max_score>=0.95] <- keep_score[input_seqs$max_score>=0.95] + 1  
  }

  if (include_virsorter) {
    keep_score[input_seqs$category==1] <- keep_score[input_seqs$category==1] + 1
    #keep_score[input_seqs$category==2] <- keep_score[input_seqs$category==2] + 0.5
    #keep_score[input_seqs$category==3] <- keep_score[input_seqs$category==3] + 0.5
    keep_score[input_seqs$category==4] <- keep_score[input_seqs$category==4] + 1
    #keep_score[input_seqs$category==5] <- keep_score[input_seqs$category==5] + 0.5 
    #keep_score[input_seqs$category==6] <- keep_score[input_seqs$category==6] + 0.5 
  }

  if (include_deepvirfinder) {
     # add if DVF calls viral
    keep_score[(input_seqs$score>=0.9 & input_seqs$pvalue<=0.05) & input_seqs$checkv_length<20000] <- keep_score[(input_seqs$score>=0.9 & input_seqs$pvalue<=0.05) & input_seqs$checkv_length<20000] + 1
   # keep_score[(input_seqs$score>=0.7 & input_seqs$pvalue<=0.05) & input_seqs$checkv_length<20000] <- keep_score[(input_seqs$score>=0.7 & input_seqs$pvalue<=0.05) & input_seqs$checkv_length<20000] + 0.5
  }

@@

this function builds a list of all of the combinations that the user wants to 
test. 
In this case, we're comparing the performance of all unique combinations of the 
six tools.
```{r}
build_score_list <- function(input_seqs, combos) {
  output <- data.frame(precision=rep(0, nrow(combos)),
                       recall=rep(0, nrow(combos)),
                       F1=rep(0, nrow(combos)),
                       MCC=rep(0, nrow(combos)),
                       prop_viral=rep(0, nrow(combos)))
  for (i in 1:nrow(combos)) {
    keep_score <- getting_viral_set_2(input_seqs, include_vibrant = combos$VIBRANT[i],
                                            include_virsorter = combos$VS[i],
                                            include_virsorter2 = combos$VS2[i],
                                            include_tuning_viral = combos$tune_viral[i],
                                            include_tuning_not_viral = combos$tune_not_viral[i],
                                            include_deepvirfinder = combos$DVF[i])

@@

## Calculate the performance of each pipeline
quantifying variability between rule sets
```{r}
accuracy_scores <- data.frame(testing_set_index=rep(0, nrow(combos_list)*5),
                      precision=rep(0, nrow(combos_list)*5),
                       recall=rep(0, nrow(combos_list)*5),
                       F1=rep(0, nrow(combos_list)*5),
                       MCC=rep(0, nrow(combos_list)*5), 
                      prop_viral=rep(0, nrow(combos_list)*5))

accuracy_scores <- cbind(testing_set_index=rep(1, nrow(combos_list)),
                              build_score_list(viruses[viruses$Index==1,], combos_list))
for (i in 2:5) {
  accuracy_scores <- rbind(accuracy_scores,
                           cbind(testing_set_index=rep(i, nrow(combos_list)),
                              build_score_list(viruses[viruses$Index==i,], combos_list)))
}
```

@@

## Calculate the performance of each pipeline

@@

@@

ggplot(accuracy_scores, aes(x=toolcombo, y=MCC, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=20),
  ) +
  xlab("Tool Combination (tv, DVF, tnv, VB, VS, VS2)") +
  ylab("MCC") +
  scale_fill_manual(name="Number of Tools",
                     values = alpha(rev(viridis(6)), 0.5)) +
  scale_color_manual(name="Number of Tools",
                     values = alpha(rev(viridis(6)), 1))
```

```{r}
max(accuracy_scores$MCC)
```
=78

@@

#comparing based on 0.5 rules or not

```{r}
#accuracy_scores_melt_no_0.5 <- accuracy_scores_melt
```

```{r}
percent_diff <- data.frame(toolcombo=accuracy_scores_melt$toolcombo,
                           percent_diff=(accuracy_scores_melt$performance_metric_score -
                   accuracy_scores_melt_no_0.5$performance_metric_score)/
  accuracy_scores_melt$performance_metric_score,
                          performance_metric=accuracy_scores_melt$performance_metric)
)

fig <- ggplot(percent_diff, 
              aes(x=performance_metric, y=percent_diff)) +
  geom_boxplot() +
  #geom_point() +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=10),
    axis.text.x=element_text(size=12),
    legend.text=element_text(size=12),
    axis.title=element_text(size=12),
    strip.background = element_rect(fill="white", color="grey"),
    strip.text = element_text(color="black", size=12)
  ) +
  xlab("Performance Metric") +
  ylab("Percent Difference") +
  geom_hline(yintercept = 0, color="grey")

```{r}
t.test(accuracy_scores_melt$performance_metric_score[accuracy_scores_melt$performance_metric=="MCC"],
       accuracy_scores_melt_no_0.5$performance_metric_score[accuracy_scores_melt_no_0.5$performance_metric=="MCC"])
```

```{r}
t.test(accuracy_scores_melt$performance_metric_score[accuracy_scores_melt$performance_metric=="precision"],
       accuracy_scores_melt_no_0.5$performance_metric_score[accuracy_scores_melt_no_0.5$performance_metric=="precision"])
```

```{r}
t.test(accuracy_scores_melt$performance_metric_score[accuracy_scores_melt$performance_metric=="recall"],
       accuracy_scores_melt_no_0.5$performance_metric_score[accuracy_scores_melt_no_0.5$performance_metric=="recall"])
```

```{r}
t.test(accuracy_scores_melt$performance_metric_score[accuracy_scores_melt$performance_metric=="prop_viral"],
       accuracy_scores_melt_no_0.5$performance_metric_score[accuracy_scores_melt_no_0.5$performance_metric=="prop_viral"])
```
